\documentclass{beamer}


\usepackage{amsmath}
\usepackage[style=alphabetic,url=true]{biblatex}
\usepackage{environ}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage[T2A]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[cache=false]{minted}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{calrsfs}
\usepackage{animate}
\usepackage{xmpmulti}


% \usetheme{Bergen}

\usecolortheme{beaver}

\setbeamertemplate{itemize item}[circle]
\setbeamertemplate{itemize subitem}{--}
\addtobeamertemplate{navigation symbols}{}{
  \usebeamerfont{footline}%
  \usebeamercolor[fg]{footline}%
  \hspace{1em}%
  \insertframenumber/\inserttotalframenumber
}
\graphicspath{ {./graphics/} }
\setminted[Python]{
  fontsize=\tiny
}
\setminted[Lisp]{
  fontsize=\tiny
}
\BeforeBeginEnvironment{minted}{\medskip}
\AfterEndEnvironment{minted}{\medskip}
\usetikzlibrary{matrix}
\tikzset{
  stack/.style={
    matrix of nodes,
    nodes={
      fill=lightgray,draw,text=black,font=\sffamily\bfseries,
      text height=11pt,text depth=3pt,baseline=center, minimum width=1cm
    },
    column sep=-\pgflinewidth/2
  }
}

\title{
  Bitcoin and Cryptocurrency Technologies \\
  Lecture 10: Other Cryptocurrencies
}

\author{Yuri Zhykin}
\date{May 30, 2021}

\begin{document}

\frame{\titlepage}

\begin{frame}
  \frametitle{Problems and Solutions}
  \begin{itemize}
  \item As of 2021, there is more than 4000 cryptocurrencies, most of which
    have no useful application.
  \item Various cryptocurrencies claim to ``solve'' one or several problems of
    Bitcoin protocol, such as:
    \begin{itemize}
    \item weak/inflexible scripting system,
    \item mining centralization,
    \item energy consumption,
    \item \textbf{transaction privacy}.
    \end{itemize}
  \item Bitcoin, unlike most other cryptocurrency systems, does not have a
    \textbf{central governing body}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Scripting System 1/3}
  \begin{itemize}
  \item One of the first ideas of the ``next generation'' cryptocurrencies: make
    the transaction validation language Turing-complete to allow arbitrary
    ``smart contracts''.
  \item A \textbf{smart contract} is a \textbf{self-executing} contract with the
    terms of the agreement between the parties specified as code.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Scripting System 2/3}
  \begin{itemize}
  \item Bitcoin Script is a \textbf{simple stack-based Turing-incomplete}
    language by design.
  \item Bitcoin's approach:
    \begin{itemize}
    \item simple language is easier to analyze for incorrect behavior
    \item Turing-incompleteness prevents infinite loops and undefined behavior
    \item most ``smart contracts'' are simple enough to be expressed in Bitcoin
      Script
    \item ``smart contracts'' that cannot be expressed in Bitcoin Script,
      can/should be implemented as second-layer protocols
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Scripting System 3/3}
  \begin{itemize}
  \item Main types of smart contracts on Ethereum platform:
    \begin{itemize}
    \item Fungible tokens similar to stocks (ICOs based on ERC-20 protocol,
      2017).
    \item Non-fungible collectible tokens like CryptoKitties (2017) or NFTs
      (2020).
    \item Digital identity management.
    \end{itemize}
  \item All of the above can be implemented on top of Bitcoin using the
    \textbf{RGB protocol} within the Lightning Network, providing both
    \textbf{contract flexibility} and \textbf{scalability}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Mining Centralization 1/2}
  \begin{itemize}
  \item Bitcoin's Proof-of-Work system uses simple SHA-256d (double SHA-256)
    algorithm.
  \item SHA-256d can be easily implemented in specialized hardware
    (\textbf{ASICs} - \textbf{Application specific integrated circuits}.
  \item As of 2021, Bitcoin is mined solely on ASICs, as any generic purpose
    hardware is way to slow to be profitable.
  \item Bitcoin ASICs are mostly produced in China, which, combined with the low
    electricity costs, resulted in an accumulation of significant amounts of
    \textbf{hash-power} in mainland China.
  \item Some people within cryptocurrency community consider ASIC resistance to
    be an important characteristic of the ``next generation'' cryptocurrency
    system.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Mining Centralization 1/2}
  \begin{itemize}
  \item Bitcoin's point of view:
    \begin{itemize}
    \item ASICs are slowly approaching theoretical efficiency limits, so there
      is less incentive to install them close to the manufacturer (i.e. in
      China),
    \item a decentralization tendency has been observed in the last several
      years - more mining operations are being set up in other countries,
    \item that said, there are discussions of a potential change of the PoW
      algorithm in a distant future.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Energy Consumption}
  \begin{itemize}
  \item Lately, Bitcoin is being aggressively criticized by the mainstream
    media for its total energy consumption.
  \item Most of the claims on Bitcoin's energy consumption show little to no
    understanding of both Bitcoin and electricity usage:
    \begin{itemize}
    \item excess electricity cannot be stored,
    \item miners are incentivized to put their equipment close to cheap excess
      electricity sources,
    \item a consumer consumes too much electricity iff it causes electrical grid
      outages in the area where consumption occurs, otherwise it's just excess
      electricity consumption,
    \item value proposition of Bitcoin far outweighs the costs of electricity
      consumed.
    \end{itemize}
  \item This has been discussed in great details by Nic Carter on Twitter.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Proof of Stake}
  \begin{itemize}
  \item \textbf{Proof of stake} (PoS) protocols are a class of consensus
    mechanisms for blockchains that work by selecting validators in proportion
    to their quantity of holdings in the associated cryptocurrency.
  \item The first functioning use of PoS for cryptocurrency was Peercoin in
    2012.
  \item Proof of stake violates the energy-based interpretation of decentralized
    consensus: \textbf{Bitcoin blocks are secured by large amounts of energy
      needed to generate one}.
  \item ``Proof of Stake is why Proof of Work was invented.'' - @notgrubles on
    Twitter.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Transaction Privacy 1/2}
  \begin{itemize}
  \item Bitcoin block chain data is transparent: for every ``chunk'' of bitcoin
    created by a coinbase transaction, the whole transaction history can be
    followed by simply looking at the chain data.
  \item Bitcoin addresses are \textbf{pseudonymous}: outputs to the same address
    can be easily tied together.
  \item As a result, Bitcoin chain is susceptible to \textbf{chain analysis}
    that in some cases allows to establish the sender and the recipient of the
    bitcoin (the amount is publicly visible directly).
  \item This makes Bitcoin hard to use as a currency for daily payments, where
    transaction privacy is essential (salary, medication bills, etc).
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Transaction Privacy 2/2}
  \begin{itemize}
  \item Lightning Network significantly improves transaction privacy by moving
    most of the transactions off chain.
  \item Channel open and channel close transactions are still visible on chain.
  \item Proposals to implement privacy features on Bitcoin \textbf{sidechains}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Case Study: Ethereum 1/2}
  \begin{itemize}
  \item Created by Vitalik Buterin - a Russian-Canadian programmer and an active
    member of Bitcoin community in the early days.
  \item Launched in 2014.
  \item Main difference from Bitcoin: Turing-complete smart contract language
    Solidity with JavaScript-like syntax.
  \item In order to avoid the infinite loops and transaction spam, Ethereum's
    fee system requires users to pay for operations executed by their smart
    contracts - this is called \textbf{paying for gas}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Case Study: Ethereum 2/2}
  \begin{itemize}
  \item Turing-completeness of the Ethereum was the reason behind the infamous
    ``DAO hack'' - loss of a large amount of funds in 2016 that caused
    \textbf{the Ethereum governing body to revert a transaction via a hard
      fork}.
  \item An investigation in 2019 indicated that 60\% of Ethereum nodes were
    running in the cloud on centralized platforms like AWS.
  \item Ethereum's scalability problem is even worse than that of Bitcoin: full
    archive is approximately 7.3 Tb of data, while the regular pruned node size
    is 300 Gb.
  \item Ethereum network is determined to move to a Proof-of-Stake system.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Case Study: Monero 1/2}
  \begin{itemize}
  \item Based on the CryptoNote protocol described by a pseudonymous entity
    \textbf{Nickolas van Saberhagen}.
  \item Created by a pseudonymous individual \textbf{thankful\_for\_today} who
    forked the codebase of CryptoNote-based coin Bytecoin and launched the
    network.
  \item Similar to Bitcoin, has no central governing body.
  \item Very simple transaction structure with no scripting system.
  \item ASIC-resistant Proof-of-Work algorithm.
  \item Variable block size, short period between blocks (2 minutes).
  \item Protocol upgrades are performed via \textbf{planned hard forks}, which
    is only feasible because the network is comparatively small.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Case Study: Monero 2/2}
  \begin{itemize}
  \item Based on the CryptoNote protocol described by a pseudonymous entity
    \textbf{Nicolas van Saberhagen}.
  \item Created by a pseudonymous individual \textbf{thankful\_for\_today} who
    forked the codebase of CryptoNote-based coin Bytecoin and launched the
    network.
  \item Has a very simple transaction structure with no scripting system.
  \item The privacy of Monero is achieved by concealing the following
    information:
    \begin{itemize}
    \item the sender of the money via the \textbf{ring signatures},
    \item the receiver of the money via the \textbf{stealth addresses},
    \item the amount sent via the Confidential Transactions system that encrypts
      the amounts in transaction but allows to check that the amount spend is
      greater than the amount received.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Conclusions}
  \begin{itemize}
  \item Bitcoin and Monero are the only cryptocurrencies whose creators are no
    longer engaging with the community.
  \item Satoshi Nakamoto never appeared online since 2011.
  \item Some people believe Satoshi disappeared to prevent any entity, including
    themselves, from manipulating the community using their name, identity,
    sentimental role, or the amount of bitcoin they presumably own.
  \item Bitcoin community is very conservative regarding the protocol changes,
    and network safety and stability are considered the most important goals.
  \item Features that Bitcoin lacks can often be implemented as layer-2
    solutions.
  \item Transaction privacy is one of the first priority goals (after safety and
    stability).
  \item \textbf{Do we need other cryptocurrencies?}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The End}
  \begin{center}
    Thank you!
  \end{center}
\end{frame}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
